Physics of Tidal Dissipation in Early-Type Stars and White Dwarfs: Hydrodynamical Simulations of Internal Gravity Wave Breaking in Stellar Envelopes

### 2020-02-23 writeups/Su_IGW_break.tex
fix tilde at last

\title[Physics of Tidal Dissipation]{Physics of Tidal Dissipation in Early-Type
Stars and White Dwarfs: Hydrodynamical Simulations of Internal Gravity Wave
Breaking in Stellar Envelopes}
\author[Y. Su et\ al.]{
Yubo Su,$^1$,
Daniel Lecoanet,$^{2,3}$
Dong Lai$^{1,4}$
\\
$^1$ Cornell Center for Astrophysics and Planetary Science, Department of
Astronomy, Cornell University, Ithaca, NY 14853, USA
\\
$^2$ Princeton Center for Theoretical Science, Princeton University, Princeton,
NJ 08544, USA
\\
$^3$ Department of Astrophysical Sciences, Princeton University, Princeton NJ
08544, USA
\\
$^4$ Tsung-Dao Lee Institute, Shanghai Jiao Tong University, Shanghai, 200240, China
}

@@

\begin{abstract}
    In binaries composed of early-type stars and white dwarfs, the dominant
    tidal process involves the excitation of internal gravity waves (IGWs),
    which propagate towards the stellar surface, and their dissipation via
    nonlinear wave breaking. We perform 2D hydrodynamical simulations of this
    wave breaking process in a stratified, isothermal atmosphere. We find that,
    after an initial transient phase, the dissipation of the IGWs naturally
    generate a sharp critical layer, separating the lower stationary region
    (with no mean flow) and the upper ``synchronized'' region (with the mean
    flow velocity equal to the horizontal wave phase speed). While the critical
    layer is steepened by these waves, it is simultaneously broadened by
    Kelvin-Helmholtz instabilities such that, in steady state, the critical
    layer width is determined by the Richardson criterion. We study the
    absorption and reflection of incident waves off the critical layer and
    provide analytical formulae describing its evolution. The result of this
    study is important for characterizing the evolution of tidally heated white
    dwarfs and other binary stars.
\end{abstract}

@@

Recent studies have identified nonlinear dissipation of IGWs as the key tidal
process in compact WD binaries \citep{fullerII,fullerIV,tidal_novae,bukart}:
IGWs are tidally excited mainly at the compositon transitions of the WD
envelope; as these waves propagate outwards towards the WD surface, they grow in
amplitude until they break, and transfer both energy and angular momentum from
the binary orbit to the outer envelope of the WD\@. However, these previous
works parameterized the wave breaking process in an ad hoc manner. The details
of dissipation, namely the location and spatial extent of the wave breaking,
affect the observable outcomes: dissipation near the surface of the WD can be
efficiently radiated away and simply brightens the WD, while dissipation deep in
the WD envelope causes an energy buildup that results in energetic flares
\citep{tidal_novae}. An important goal of this paper is to elucidate the details
of the nonlinear IGW breaking process; the result of this ``microphysics'' study
will help determine the thermal evolution and and the observational
manifestations of tidally heated binary WDs.

@@

A full video of our simulation with $N_x = 768$, $N_z = 3072$, $\mathrm{Re} =
1024$ is available
online\footnote{http://www.princeton.edu/\~lecoanet/data/breaking\_wave.mov}. We
take this to be our fiducial simulation for the remainder of this paper, though
other simulations show qualitatively similar behavior.

@@

In this paper, we have studied the nonlinear behavior of IGWs with $\abs{k_x /
k_z} \sim 1 / (2\pi)$ in a plane-parallel geometry. Tidally excited IGWs in
binary stars have horizontal wave number $k_\perp \sim 1/R$ (where $R$ is the
stellar radius) much smaller than the radial wave number $k_r$. While our
simulations do not have $\abs{k_x/k_z}\ll 1$, the qualitative behavior is likely
to be similar, as the turbulence driving the critical layer dynamics occurs at
scales significantly smaller than either $1/k_x$ or $1/\abs{k_z}$. Simulating
IGWs with $k_x \ll \abs{k_z}$ is more challenging numerically and we defer its
exploration to future work.

>>>>

\title[Physics of Tidal Dissipation]{Physics of Tidal Dissipation in Early-Type
Stars and White Dwarfs: Hydrodynamical Simulations of Internal Gravity Wave
Breaking in Stellar Envelopes}
\author[Y. Su et\ al.]{
Yubo Su,$^1$,
Daniel Lecoanet,$^{2,3}$
Dong Lai$^{1,4, 5}$
\\
$^1$ Cornell Center for Astrophysics and Planetary Science, Department of
Astronomy, Cornell University, Ithaca, NY 14853, USA
\\
$^2$ Princeton Center for Theoretical Science, Princeton University, Princeton,
NJ 08544, USA
\\
$^3$ Department of Astrophysical Sciences, Princeton University, Princeton NJ
08544, USA
\\
$^4$ Tsung-Dao Lee Institute, Shanghai Jiao Tong University, Shanghai, 200240,
China
\\
$^5$ Department of Astronomy and Miller Institute for Basic Research In Science,
UC Berkeley, Berkeley, CA 94720, USA
}

@@

\begin{abstract}
    In binaries composed of early-type stars and white dwarfs, the dominant
    tidal process involves the excitation of internal gravity waves (IGWs),
    which propagate towards the stellar surface, and their dissipation via
    nonlinear wave breaking. We perform 2D hydrodynamical simulations of this
    wave breaking process in a stratified, isothermal atmosphere. We find that,
    after an initial transient phase, the dissipation of the IGWs naturally
    generate a sharp critical layer, separating the lower stationary region
    (with no mean flow) and the upper ``synchronized'' region (with the mean
    flow velocity equal to the horizontal wave phase speed). While the critical
    layer is steepened by absorption of these waves, it is simultaneously
    broadened by Kelvin-Helmholtz instabilities such that, in steady state, the
    critical layer width is determined by the Richardson criterion. We study the
    absorption and reflection of incident waves off the critical layer and
    provide analytical formulae describing its long-term evolution. The result
    of this study is important for characterizing the evolution of tidally
    heated white dwarfs and other binary stars.
\end{abstract}

@@

Recent studies have identified nonlinear dissipation of IGWs as the key tidal
process in compact WD binaries \citep{fullerII,fullerIV,tidal_novae,bukart}:
IGWs are tidally excited mainly at the compositon transitions of the WD
envelope; as these waves propagate outwards towards the WD surface, they grow in
amplitude until they break, and transfer both energy and angular momentum from
the binary orbit to the outer envelope of the WD\@. However, these previous
works parameterized the wave breaking process in an ad hoc manner. The details
of dissipation, namely the location and spatial extent of the wave breaking,
affect the observable outcomes: dissipation near the surface of the WD can be
efficiently radiated away and simply brightens the WD, while dissipation deep in
the WD envelope causes an energy buildup that results in energetic flares
\citep{tidal_novae}. An important goal of this paper is to elucidate the details
of the nonlinear IGW breaking process; the result of this ``microphysics'' study
will help determine the thermal evolution and the observational manifestations
of tidally heated binary WDs.

@@

A full video of our simulation with $N_x = 768$, $N_z = 3072$, $\mathrm{Re} =
1024$ is available
online\footnote{\url{http://www.princeton.edu/\~lecoanet/data/breaking\_wave.mov}}. We
take this to be our fiducial simulation for the remainder of this paper, though
other simulations show qualitatively similar behavior.

@@

In this paper, we have studied the nonlinear behavior of IGWs with $\abs{k_x /
k_z} \sim 1 / (2\pi)$ in a plane-parallel geometry. Tidally excited IGWs in
binary stars have horizontal wave number $k_\perp \sim 1/R$ (where $R$ is the
stellar radius) much smaller than the radial wave number $k_r$. While our
simulations do not satisfy $\abs{k_x/k_z}\ll 1$, the qualitative behavior is likely
to be similar, as the turbulence driving the critical layer dynamics occurs at
scales significantly smaller than either $1/k_x$ or $1/\abs{k_z}$. Simulating
IGWs with $k_x \ll \abs{k_z}$ is more challenging numerically and we defer its
exploration to future work.
